A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2015-10-05 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Improve ND paper ina major way

\usepackage{fancybox}
\usepackage{latexsym}
\usepackage{proof}
\usepackage{bussproofs}
\usepackage{amsmath}
\EnableBpAbbreviations
\newcommand{\rl}[1]{\RightLabel{#1}}

@@

@@

@@

\author{Annika Siders\inst{1} \and Bruno Woltzenlogel Paleo\inst{2}}

\authorrunning{A.\~Siders \and B.\~Woltzenlogel Paleo}

\institute{
  Helsinki, Finland\\
  \email{annika.siders@helsinki.fi}
  \and 
  Vienna, Austria ; Canberra, Australia \\
  \email{bruno.wp@gmail.com}
}

@@

@@

%spell check

%Extra:

%Difference in the formalizations of Leibniz and Gödel? Why did Gödel change to conjunction?

%Kant argued against the ontological argument on the basis that existence is not an analytic property \citep{kant}. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. The strongest claim that can be proven by the ontological argument is a conditional claim of necessary existence: if God exists, then God necessarily exists \citep{citation needed}.

% Another, possibly undesirable, consequence of the formalism is that there can only be one God, which is defined by the axioms. Namely, if one assumes that the theory contains the equality relation then a God has the property of being identical to itself. Since any other god-like being would have at least the same properties it shares the property of being identical to the one God. Thus, we have proved monotheism \citep{sobel2}[Ch. 4, section 3.3.2]. %\citep{fitting}[The proof is Exercise 7.1, p.163].  Inclusion of identity relation gives monothemism?*** Reference missing?

% The question arises what kind of second order quantification is allowed. The instantiations that are used in our proof are instantiated with predicate variables that later work as eigenvariables for the second order universal introduction. In the proof of corollary \ref{C1} we instantiate with $G$ and in the proof of lemma \ref{L1}. we instantiate with $G$ and $E$. %In Sobel on G\"{o}del's ontological proof by Robert E. Koons this question is raised.

% One can also discuss if a second order modal logic is sound and complete. With Henkin models, which restrict the second order quantification with comprehension, we have that normal second order logic is sound and complete. There is a completeness proof for a second order modal logic found in A completeness theorem in second order modal logic by Cocchiarella.

% ToDo: check if disjunction rules are used in Goedel's proof. If not, we could define disjunction as we already do for negation and equivalence.

@@

Leibniz identified the possible existence of God as a critical missing step in St. Anselm's argument. To fill this gap, he argued that the properties of God, the perfections, are compatible. This means that it is possible to satisfy all perfections at once which implies that the existence of a greatest conceivable being with all these properties is possible.

G\"odel bulit on Leibniz's work \cite{adams} and brought the ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being that has all positive properties. G\"odel's work was saved in his own notes \cite{Goedel} as well as notes by Scott \cite{Scott}, in whom he confided his proof.

The increase in formality of the ontological argument has required a development of the basic notions. G\"odel's notion of positive property and Leibniz's notion of perfection differ. A formal distinction is that Leibniz's perfections are atomic whereas G\"odel's positive properties can consist of combinations of atomic properties \citep{fitting}[p.139]. In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being god-like is positive. Intuitively, a (possibly infinite) conjunction of positive properties is deduced from the universal definition of God-likeness. This deductive inference is not formalizable in a finite first-order calculus. The interplay between universal quantification (in the definition of a god-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) could explain why, starting with Scott \cite{Scott}, this axiom o!
 f G\"odel has been replaced by another that simply assumes the positivity of the property of being god-like.

The main criticism against the ontological argument is that its soundness has been doubted. As Fitting \citep{fitting}[p.134] described the problem ''we do not know whether the concept of greatest conceivable being in fact designates anything or if it is inconsistent, like a round square''.

The main criticism against G\"odel's formal argument is an undesirable consequence of the stipulated axioms, called \emph{modal collapse}. Many recent works on the ontological argument have focused on proposing modifications of the argument that would not entail a modal collapse. This is discussed in greater detail in Section \ref{sec:collapse}, but possible solutions are not implemented in this paper.

The aim of this paper is to present two detailed formal proofs of G\"odel's ontological argument in a natural deduction calculus (as defined in Section \ref{sec:calculus}). For a comprehensive introduction to natural deduction the reader can consult \citep{prawitz}. The natural deduction style was chosen for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, the natural deduction formalizations have been easily checked step-by-step in Coq \cite{coqformalizations}.

The first contribution of this paper is a detailed formalization of Scott's version \cite{Scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in a natural deduction calculus. In section \ref{sec:compl-sound} we prove that the calculus used is sound and complete relative to an axiomatic modal calculus and this detailed formalization of Scott's version has as been formally verified. The second contribution of the paper is a new proof (also in natural deduction style) in a natural deduction calculus. In contrast to Scott's proof \cite{Scott} formalized in the modal calculus S5, the latter proof is formalized in KB, as discussed in Section \ref{sec:AxiomB}.

@@

The natural deduction calculus used here has standard rules for propositional connectives and quantifiers, as shown in Figures \ref{fig:PropositionalRules} and \ref{fig:QuantifierRules}. The extension to classical logic is achieved by adding a rule for double negation elimination, shown in Figure \ref{fig:Classical}. Finally, modal operators are handled by special rules that insert or remove formulas from boxes, as shown in Figure \ref{fig:NDK}. Apart from the use of labels and the dual rules for `$\pos$', these rules are essentially the modal rules from \cite{todo}. Beta-reduction is implicit; all rules are assumed modulo beta-reduction. A \emph{derivation} is a directed acyclic graph whose nodes are formulas and whose edges correspond to applications of the inference rules. A \emph{proof} of a formula $F$ is a derivation without open assumptions and having $F$ as root not inside any box.

Double lines are used to abbreviate tedious propositional reasoning steps in the derivations. Dashed lines are used to refer to an axiom or theorem with proof shown elsewhere. Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

@@

%\clearpage

\noindent
The calculus having only the rules shown in Figures \ref{fig:PropositionalRules}, \ref{fig:Classical} and \ref{fig:QuantifierRules} is named \ND. The calculus with the additional rules shown in Figure \ref{fig:NDK} is called \NDK.

%\clearpage

@@

Without the restriction that every box must be accessed by exactly one strong modal inference, the calculus would be unsound for the modal logic \textbf{K}, because the formula $\all \psi. (\nec \psi \imp \pos \psi)$ would be derivable although it is not valid in \textbf{K}.

@@

This example of an unsound derivation is prevented by the eigen-box condition because the box labelled by $\omega$ is not accessed by any strong inference.

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible worlds} semantics, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy has already been contested \cite{fitting}.

\section{Derivable Modal Principles}
\label{sec:AxiomB}.

In standard first-order modal logics the modal axioms correspond to geometric frame properties for Kripke models \cite{negri}. The modal axiom {\bf T} corresponds to reflexivity of the frame and {\bf B} corresponds to symmetry of the frame. The frames of the modal logic {\bf S5} are in addition transitive.

There are discussion of what modal principles are needed for the ontological argument. 
G\"odel mentions in his notes that the modal system required for proving his theorem is a second order version of the modal system {\bf S5}. In \cite{and90} it is conjectured that the principles of proof could be restricted to the weaker modal logic {\bf KB}. The modal logic {\bf KB} consists of the basic modal system {\bf K} and the modal axioms {\bf T} and {\bf B}.

Through an analysis of the proofs weaker systems, in particular KB, can be shown to be sufficient for versions of the ontological argument \citep{sobel2}[p. 152]. The proof presented in Section \ref{sec:newproof} is such a version and explicitly shows a single use of Brouwer Reduction Principle, which, in turn, is provable from axiom B. All other proof steps are formalizable in a basic second-order modal logic K.

We will prove the necessary modal principles as two lemmas; a modal distribution principle derivable in K and Brouwer's reduction principle derivable in KB. In addition we prove an iteration principle, which is valid in S5, for the proof in section \ref{sec:scott}. The iteration principle is derived in KB with axiom 5.

\subsection{A modal distribution principle in K}

\begin{lemma}\label{DP}
The distribution principle 
$$\nec (A\imp B)\imp(\pos A\imp \pos B)$$
is provable in our system K of modal logic. 
\end{lemma}

@@

\subsection{Brouwer's Reduction Principle}

\begin{lemma}\label{BRP}
The Brouwer Reduction Principle $\pos\nec A\imp A$ is derivable from axiom B, $A\imp \nec \pos A$. 
\end{lemma}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $}\RightLabel{$2$}
\UIC{$\neg\nec \neg \nec A$}
\AXC{$ $}\RightLabel{$ 1$}
\UIC{$\neg A$}
\AXC{Axiom B}\dashedLine
\UIC{$\neg A\imp \nec\neg\nec\neg\neg A$}\RightLabel{$\imp_E$}
\BIC{$\nec\neg\nec\neg\neg A$}\doubleLine
\UIC{$\nec\neg\nec A$}
\BIC{$(\neg\nec \neg \nec A)\&(\nec\neg \nec A)$}\RightLabel{$\neg_I^1$}
\UIC{$\neg\neg A$} \RightLabel{$\neg\neg_E$}
\UIC{$A$} \RightLabel{$\imp_I^2$}
\UIC{$\neg\nec \neg \nec A\imp A$}
\end{prooftree}
\end{small}

@@

\subsection{S5 and its Iteration Principle}

\begin{lemma}
What is possibly necessary is necessary:
$$
\pos \nec A \imp \nec A
$$
\end{lemma}

Axiom 5 is $\pos A\imp \nec \pos A$.

\AXC{Brouwer's reduction principle}\dashedLine
\UIC{$\pos \nec A\imp A$}\RightLabel{$\nec_I$}
\UIC{$\nec(\pos \nec A\imp A)$}
\AXC{Axiom K}\dashedLine
\UIC{$\nec(\pos \nec A\imp A)\imp (\nec\pos \nec A\imp \nec A)$}\RightLabel{$\imp_E$}
\BIC{$\nec\pos \nec A\imp \nec A$}

\AXC{Axiom 5 on $\nec A$}\dashedLine
\UIC{$\pos \nec A\imp \nec \pos \nec A$}

\AXC{$ $}\RightLabel{$1$}
\UIC{$\pos \nec A$}\RightLabel{$\imp_E$}
\BIC{$\nec \pos \nec A$}\RightLabel{$\imp_E$}
\BIC{$\nec A$}\RightLabel{$\imp_I^1$}
\UIC{$\pos \nec A \imp  \nec A$}
\end{prooftree}
\end{small}

@@

The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The following proof is a detailed formalization of Scott's proof that has been verified in Coq \cite{***}. The formal argument has  two parts; a proof that if the exemplification of the god-likeness property is possible, then it is necessary and a proof that the exemplification is in fact possible.

@@

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
P(\varphi) \imp \pos \ex x.\varphi(x)
$$
\end{theorem}
\begin{proof} \hfill

\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
		\AXC{Reflexivity} \dashedLine
		\UIC{$\omega: \gamma = \gamma$} \RightLabel{$\imp_I$}
		\UIC{$\omega: \rho(\gamma) \imp \gamma = \gamma$} \RightLabel{$\all_I$}
		\UIC{$\omega: \all x. \rho(x) \imp x = x$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\all x. \rho(x) \imp x = x) $} \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x = x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x = x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x = x) \imp P(\lambda x. x = x)$}

		\BIC{$P(\lambda x. x = x)$} 
\end{prooftree}

\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
	\AXC{$ $} \RightLabel{1}
	\UIC{$\omega: \rho(\beta)$}
			\AXC{$ $} \RightLabel{3}
			\UIC{$\nec \all x. \neg \rho(x)$} \RightLabel{$\nec_E$}
			\UIC{$\omega: \all x. \neg \rho(x) $} \RightLabel{$\all_E$}
			\UIC{$\omega: \neg \rho(\beta)$} \RightLabel{$\neg_E$}
		\BIC{$\omega: \bot$} \RightLabel{$\neg_I^9$}
		\UIC{$\omega: \neg (\beta = \beta)$} \dottedLine
		\UIC{$\omega: (\beta \neq \beta)$} \RightLabel{$\imp_I^1$}
		\UIC{$\omega: (\rho(\beta) \imp (\beta \neq \beta))$} \RightLabel{$\all_I$}
		\UIC{$\omega: (\all x. \rho(x) \imp (x \neq x))$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\all x. \rho(x) \imp (x \neq x))$}  \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x \neq x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x \neq x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x \neq x) \imp P(\lambda x. x \neq x)$} 
		\BIC{$P(\lambda x. x \neq x)$} 
\end{prooftree}

\begin{prooftree}
		\AXC{$ $} \dashedLine
		\UIC{$P(\lambda x. x \neq x)$}    
    		\AXC{$ $} \dashedLine
			\UIC{$P(\lambda x. x = x)$} 
					\AXC{Half of Axiom 1} \dashedLine
					\UIC{$P(\lambda x. x = x) \imp \neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
				\BIC{$\neg P(\lambda x. x \neq x)$} \RightLabel{$\neg_E$}
			\BIC{$\bot$}
\end{prooftree}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{fancybox}
\usepackage{latexsym}

\usepackage{proof}
\usepackage{bussproofs}
\EnableBpAbbreviations
\newcommand{\rl}[1]{\RightLabel{#1}}

\usepackage{amsmath}
\usepackage{ntheorem}

\usepackage[margin=3.5cm]{geometry}

@@

\pagestyle{plain}

@@

\newtheorem*{lemma*}{Lemma}

@@

% ToDo: uncomment this post-acceptance: the journal uses double blind reviewing.
% \author{Annika Siders\inst{1} \and Bruno Woltzenlogel Paleo\inst{2}}

% \authorrunning{A.\~Siders \and B.\~Woltzenlogel Paleo}

% \institute{
%   Helsinki, Finland\\
%   \email{annika.siders@helsinki.fi}
%   \and 
%   Vienna University of Technology, Austria ; Australian National University \\
%   \email{bruno.wp@gmail.com} 
% }

\author{}
\authorrunning{}
\institute{}

@@

\begin{abstract}
This paper presents two detailed formalizations of ontological arguments in a simple natural deduction calculus. The first formal proof closely follows the hints in Scott's manuscript about G\"odel's argument and fills its gaps, thus verifying its correctness. The second formal proof improves the first one, by: firstly, relying on the weaker modal logic {\bf KB} instead of the stronger modal logic {\bf S5} in the proof of the last theorem;  and, secondly, not using equality to prove the first theorem. The second proof is also technically shorter than the first one, because it avoids unnecessary detours and uses Axiom 1 only once.
\end{abstract}

\noindent
\textbf{Keywords: } G\"odel's Ontological Proof, Higher-Order Modal Logics, Natural Deduction

@@

@@

Leibniz identified the possible existence of God as a critical missing step in St. Anselm's argument. To fill this gap, he argued that the properties of God, the perfections, are compatible. This means that it is possible to satisfy all perfections at once, which implies that the existence of a greatest conceivable being with all these properties is possible.

G\"odel bulit on Leibniz's work \cite{adams} and brought the ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being that has all positive properties. G\"odel's work was saved in his own notes \cite{Goedel} as well as notes by Scott \cite{scott}, in whom he confided his proof.

The increase in formality of the ontological argument has required a development of its basic notions. G\"odel's notion of positive property and Leibniz's notion of perfection differ. A formal distinction is that Leibniz's perfections are atomic whereas G\"odel's positive properties can consist of combinations of atomic properties \citep{fitting}[p.139]. In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being god-like is positive. Intuitively, a (possibly infinite) conjunction of positive properties is deduced from the universal definition of God-likeness. This deductive inference is not formalizable in a finite first-order calculus. The interplay between universal quantification (in the definition of a god-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) could explain why, starting with Scott \cite{scott}, this axiom of G\"odel has been replaced by another that simply assumes the positivity of the property of being god-like.

The aim of this paper is to present two detailed formalizations of ontological arguments in a natural deduction calculus. For a comprehensive introduction to natural deduction, the reader can consult \citep{prawitz}. The higher-order modal natural deduction calculus proposed here has special introduction and elimination rules for modalities, as defined in Section \ref{sec:calculus}, and is sound and complete relative to an axiomatic modal calculus, as proven in Section \ref{sec:compl-sound}.

The natural deduction style was chosen for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of proof calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, natural deduction formalizations can be verified step-by-step in such proof assistants, and we have in fact done this.

The first contribution of this paper is a detailed formalization of Scott's version \cite{scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in the proposed natural deduction calculus. The second contribution of the paper is a new proof (also in natural deduction style). In contrast to Scott's proof \cite{scott}, which requires the modal logic \textbf{S5}, the new proof requires only the weaker modal logic \textbf{KB}. The new proof also does not rely on the equality predicate and is much shorter.

A major criticism against G\"odel's formal argument is an undesirable consequence of the stipulated axioms, called \emph{modal collapse}. This is discussed in greater detail in Section \ref{sec:collapse}, where a natural deduction derivation of the collapse is presented. Although many recent works on the ontological argument have proposed modifications of the argument that do not entail a modal collapse, these solutions are beyond the scope of this paper.

@@

The natural deduction calculus used here has standard rules for propositional connectives and quantifiers, as shown in Figures \ref{fig:PropositionalRules} and \ref{fig:QuantifierRules}. The extension to classical logic is achieved by adding a rule for double negation elimination, shown in Figure \ref{fig:Classical}. Finally, modal operators are handled by special rules that insert or remove formulas from boxes, as shown in Figure \ref{fig:NDK}.
%Apart from the use of labels and the dual rules for `$\pos$', these rules are essentially the modal rules from \cite{todo}. 
Beta-reduction is implicit; all rules are assumed to operate modulo beta-reduction. A \emph{derivation} is a directed acyclic graph whose nodes are formulas and whose edges correspond to applications of the inference rules. A \emph{proof} of a formula $F$ is a derivation without open assumptions and having $F$ as root not inside any box.

\newcommand{\subproof}{\star}

Double lines are used to abbreviate tedious propositional reasoning steps in the derivations. Dashed lines are used to refer to an axiom or theorem with proof shown elsewhere. When proof trees are too large to fit in the page, some branches may be displayed separately. In such cases, the conclusion of the branch and the location in the main proof tree where the branch belongs are annotated with the same symbol (a subscripted $\subproof$). Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

The calculus having only the rules shown in Figures \ref{fig:PropositionalRules}, \ref{fig:Classical} and \ref{fig:QuantifierRules} is named \ND, while the calculus with the additional rules shown in Figure \ref{fig:NDK} is named \NDK.

@@

@@

Without the restriction that every box must be accessed by exactly one strong modal inference, the calculus would be unsound for the modal logic \textbf{K}. The formula $\all \psi. (\nec \psi \imp \pos \psi)$, for example, is not valid in \textbf{K} but would be provable according to the following proof:

@@

The example proof above is unsound according to the {\NDK} calculus, because the eigen-box condition is violated: the box labelled by $\omega$ is not accessed by any strong inference.

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible worlds} semantics, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \cite{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \cite{hajek}). Nevertheless, some authors of variants \cite[footnotes 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of assuming constant domains is motivated by simplicity.

\section{Some Useful Derivable Modal Principles}
\label{sec:ModalLogics}

From an axiomatic point of view, modal logics differ with respect to which additional axiom schemas they admit. Some common axiom schemas of relevance to the ontological argument are T ($\nec A \imp A$), B ($A \imp \nec \pos A$), 4 ($\nec A \imp \nec \nec A $) and 5 ($\pos A \imp \nec \pos A$). From a semantic point of view, these axioms correspond to geometric frame conditions that must be satisfied by the accesibility relation in the Kripke models \cite{negri}: T corresponds to reflexivity, B corresponds to symmetry, 4 corresponds to transitivity and 5 corresponds to euclidianity. In the models of logic {\bf S5}, all these axioms are satisfied; in the models of logic {\bf KB}, K and B are satisfied; and in the models of logic {\bf B}, K, T and B are satisfied.

In this section, natural deduction proofs of a three convenient derived modal principles are presented. These principles are used in the ontological proofs formalized in the next sections. Although they are well-known among modal logicians, we include their proofs here in order to have a self-contained and fully formal presentation of the ontological proofs, without any gaps.

\bigskip

\noindent
The \emph{distribution principle} can be seen as a form of modus ponens within the scope of modalities: if $A \imp B$ holds in all accessible worlds and $A$ holds in an accessible world, then $B$ holds in an accessible world. This principle is provable in the modal logic {\bf K}.

\begin{lemma*}[Distribution Principle]
\label{DP} 
$$\nec (A\imp B)\imp(\pos A\imp \pos B)$$
\end{lemma*}

@@

\bigskip

\noindent
\emph{Brouwer's reduction principle} is derivable in modal logic {\bf KB}, using axiom B. The proof of G\"odel and Scott (Section \ref{sec:scott}) do not make direct use of this principle, but the new proof presented in Section \ref{sec:newproof} does.

\begin{lemma*}[Brouwer's Reduction Principle]
\label{BRP}
$$
\pos\nec A\imp A
$$ 
\end{lemma*}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $}\RightLabel{$2$}
\UIC{$\pos \nec A $} \doubleLine
\UIC{$\neg\nec \neg \nec A$}
\AXC{$ $}\RightLabel{$ 1$}
\UIC{$\neg A$}
\AXC{Axiom B}\dashedLine
\UIC{$\neg A\imp \nec\pos\neg A$} \doubleLine
\UIC{$\neg A\imp \nec\neg\nec\neg\neg A$}\RightLabel{$\imp_E$}
\BIC{$\nec\neg\nec\neg\neg A$}\doubleLine
\UIC{$\nec\neg\nec A$} \RightLabel{$\neg_E$}
\BIC{$\bot$}\RightLabel{$\neg_I^1$}
\UIC{$\neg\neg A$} \RightLabel{$\neg\neg_E$}
\UIC{$A$} \RightLabel{$\imp_I^2$}
\UIC{$\pos \nec A\imp A$}
\end{prooftree}
\end{small}

@@

\bigskip

\noindent
In the modal logic {\bf S5}, a sequential iteration of modalities can be collapsed to the last modality in the sequence. In other words, the following principles hold in {\bf S5}:
\begin{itemize}
\item $\pos^n \nec A \imp \nec A$
\item $\nec^n \pos A \imp \pos A$
\item $\pos^n \pos A \imp \pos A$
\item $\nec^n \nec A \imp \nec A$
\end{itemize}

\noindent
Below we exhibit a natural deduction proof for the first principle listed above  when $n = 1$. Informally, this particular case can be read as the claim that anything that is possibly necessary is in fact necessary, and it is sufficient for the ontological proofs of G\"odel and Scott. The proof of this principle depends on the previously proven Brouwer's reduction principle, on the Modal Axiom 5, and on K, whose derivation is shown in the proof of Theorem \ref{theorem:completeness}.

\begin{lemma*}[Iteration Principle]
$$
\pos \nec A \imp \nec A
$$
\end{lemma*}

\AXC{Brouwer's Reduction}\dashedLine
\UIC{$\pos \nec A\imp A$}\RightLabel{$\nec_I$}
\UIC{$\nec(\pos \nec A\imp A)$}
\AXC{K}\dashedLine
\UIC{$\nec(\pos \nec A\imp A)\imp (\nec\pos \nec A\imp \nec A)$}\RightLabel{$\imp_E$}
\BIC{$\nec\pos \nec A\imp \nec A$}

\AXC{Axiom 5 for $\nec A$}\dashedLine
\UIC{$\pos \nec A\imp \nec \pos \nec A$}

\AXC{$ $}\RightLabel{$1$}
\UIC{$\pos \nec A$}\RightLabel{$\imp_E$}
\BIC{$\nec \pos \nec A$}\RightLabel{$\imp_E$}
\BIC{$\nec A$}\RightLabel{$\imp_I^1$}
\UIC{$\pos \nec A \imp  \nec A$}
\end{prooftree}
\end{footnotesize}

@@

\newcommand{\scott}{^{\dagger}}

In this section we present a detailed formalization of Scott's proof in the natural deduction calculus defined in Section \ref{sec:calculus}. The reasoning in Scott's manuscript was reproduced step-by-step and any reasoning gaps were filled in ways that we considered most natural. All derived formulas that do appear in Scott's manuscript are marked with a $\dagger$ here. Unmarked formulas were derived in the process of filling the gaps between the marked formulas mentioned in Scott's manuscript.

% ToDo: uncomment this post-acceptance: the journal uses double blind reviewing.
%The formal natural deduction proof presented here was verified in the Coq proof assistant\footnote{\url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Coq/GoedelGod_Scott.v}}.

Scott's version of G\"odel's proof depends on 5 axioms that circumscribe the notion of \emph{positive} property, with positivity being denoted by the undefined second-order predicate symbol $P$. Additionally, 3 definitions are used for defining the notions of \emph{God-like}, \emph{essence} and \emph{necessary existence}. Technically, these definitions only abbreviate certain complex formulas. The argument would still go through if all defined symbols were replaced by the complex formulas they define. This observation is particularly relevant in the case of \emph{necessary existence}. Since this notion of ``existence'' is just an abbreviation, G\"odel's argument is not susceptible to Kant's criticism against Anselm's argument (that existence should not be treated as a predicate). In G\"odel's proof, existence is properly denoted by the existential quantifier. It is, therefore, unfortunately misleading to refer to the defined predicate symbol $E$ as ``necessary existence'', when in fact it is just a convenient abbreviation.

Besides the 5 axioms and 3 definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the equality axiom of reflexivity (in the proof of Theorem 1). The use of reflexivity could be avoided, but it gives a shorter proof.

@@

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}
\begin{proof} \hfill

\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
		\AXC{Reflexivity} \dashedLine
		\UIC{$\omega: \gamma = \gamma$} \RightLabel{$\imp_I$}
		\UIC{$\omega: \rho(\gamma) \imp \gamma = \gamma$} \RightLabel{$\all_I$}
		\UIC{$\omega: \all x. \rho(x) \imp x = x$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\all x. \rho(x) \imp x = x)\scott $} \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x = x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x = x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x = x) \imp P(\lambda x. x = x)$} 
		\BIC{$[\subproof_1] \qquad P(\lambda x. x = x)\scott$} 
\end{prooftree}

\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
	\AXC{$ $} \RightLabel{1}
	\UIC{$\omega: \rho(\beta)$}
			\AXC{$ $} \RightLabel{3}
			\UIC{$\nec \all x. \neg \rho(x)$} \RightLabel{$\nec_E$}
			\UIC{$\omega: \all x. \neg \rho(x) $} \RightLabel{$\all_E$}
			\UIC{$\omega: \neg \rho(\beta)$} \RightLabel{$\neg_E$}
		\BIC{$\omega: \bot$} \RightLabel{$\bot_E$}
		\UIC{$\omega: \neg (\beta = \beta)$} \dottedLine
		\UIC{$\omega: (\beta \neq \beta)$} \RightLabel{$\imp_I^1$}
		\UIC{$\omega: (\rho(\beta) \imp (\beta \neq \beta))$} \RightLabel{$\all_I$}
		\UIC{$\omega: (\all x. \rho(x) \imp (x \neq x))$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\all x. \rho(x) \imp (x \neq x))\scott$}  \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x \neq x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x \neq x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x \neq x) \imp P(\lambda x. x \neq x)$} 
		\BIC{$[\subproof_2] \qquad P(\lambda x. x \neq x)\scott$} 
\end{prooftree}

\begin{prooftree}
		\AXC{$\subproof_2$} \dashedLine
		\UIC{$P(\lambda x. x \neq x)\scott$}    
    		\AXC{$\subproof_1$} \dashedLine
			  \UIC{$P(\lambda x. x = x)\scott$} 
					\AXC{Half of Axiom 1} \dashedLine
					\UIC{$P(\lambda x. x = x) \imp \neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
				\BIC{$\neg P(\lambda x. x \neq x)$} \RightLabel{$\neg_E$}
			\BIC{$\bot$}
\end{prooftree}

\begin{prooftree}
\AXC{$ $} \RightLabel{4}
\UIC{$\nec \neg \ex x. \rho(x)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x. \rho(x)$}
        \AXC{$ $} \RightLabel{2}
        \UIC{$\rho(\alpha)$} \RightLabel{$\ex_I$}
		\UIC{$\omega: \ex x. \rho(x)$} \RightLabel{$\neg_E$}
    \BIC{$\omega: \bot$} \RightLabel{$\neg_I^2$}
	\UIC{$\omega: \neg \rho(\alpha)$} \RightLabel{$\all_I$}
	\UIC{$\omega: \all x. \neg \rho(x)$} \RightLabel{$\nec_I$}
	\UIC{$\nec \all x. \neg \rho(x) \scott$}
	        \AXC{$ $} \dashedLine
	        \UIC{$\bot$} \RightLabel{$\neg_I^3$}
    		\UIC{$\neg \nec \all x. \neg \rho(x)$} \RightLabel{$\neg_E$}
  		\BIC{$\bot$} \RightLabel{$\neg_I^4$}
  		\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
  		\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^5$}
  		\UIC{$ P(\rho) \imp \pos \ex x.\rho(x)$} \RightLabel{$\all_I$}
      \UIC{$ \all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)] $}
\end{prooftree}
